---
title: 'Final Project - Step 2 (15 Points)'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author:
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```



:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Xinyao Song   (xinyaosong)
- Matthew Chan (hchan837)
- Darren Colianni    (dcolianni)
- STUDENT 4    (NetID 4)
- STUDENT 5    (NetID 5)
:::


::: {.callout-caution}
## Due Date
The deadline for this step is **Friday, May 9, 2025**.
:::


:::{.callout-tip}
# Instructions
In this step, you will develop clear research questions and hypotheses based on your selected dataset, and conduct a thorough Exploratory Data Analysis (EDA). This foundational work is crucial for guiding your analysis in the following steps.

:::

# Step 2: Research Questions, Hypotheses, and Exploratory Data Analysis (EDA)


## Research Questions

**Question 1**  

- How does weekly study time, parents’ education level, and internet access influence a student’s average Portuguese grade (average of G1, G2, and G3)?

**Question 2**  

- How is alcohol consumption among Portuguese high school students associated with their academic performance and social activities? 

**Question 3**  

- Does the average free time differ between students who are in a romantic relationship and those who are not? 

## Hypotheses

**Hypothesis 1**  

- At least one of the predictors (weekly study time, parents’ education level, or internet access) has a significant effect on students’ average Portuguese grade. 

**Hypothesis 2**  

- Alcohol consumption is significantly associated with either academic performance or social activity levels among Portuguese high school students. 

**Hypothesis 3**  

- There is a significant difference in means between students in a relationship and those who are not.

## Exploratory Data Analysis (EDA)

## Data Cleaning
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)

# Preprocess student data frame
por_df <- readr::read_delim("../data/student-por.csv", delim = ";")

por_df <- por_df %>%
  mutate(
    school = as.factor(school),
    sex = as.factor(sex),
    address = as.factor(address),
    famsize = as.factor(famsize),
    Pstatus = as.factor(Pstatus),
    Mjob = as.factor(Mjob),
    Fjob = as.factor(Fjob),
    reason = as.factor(reason),
    guardian = as.factor(guardian),
    schoolsup = as.factor(schoolsup),
    famsup = as.factor(famsup),
    paid = as.factor(paid),
    activities = as.factor(activities),
    nursery = as.factor(nursery),
    higher = as.factor(higher),
    internet = as.factor(internet),
    romantic = as.factor(romantic),
    # ordered factor from 1-5 for very high, low etc.
    Dalc = factor(Dalc, levels = 1:5, ordered = TRUE), 
    Walc = factor(Walc, levels = 1:5, ordered = TRUE)
  )

# Check for missing values
print(anyNA(por_df))
```
From the missing value check above, we can see that there are no missing values in this dataset. Therefore, no data cleaning is necessary.

## Descriptive Statistics
```{r}
summary(por_df)
```



## Data Visualization

### Research Question 3: Impact of Non-previous grade factors on grade

```{r, warning=FALSE, echo=FALSE}
library(ggplot2)
library(patchwork)

make_boxplot <- function(por_df, xvar, fill_colors = NULL,  # boxplot gg func
                         title = NULL, xlabel = NULL, ylabel = "G3") {
  p <- ggplot(df, aes_string(x = xvar, y = "G3", fill = xvar)) +
    geom_boxplot() +
    labs(
      title = ifelse(is.null(title), paste(xvar, "vs Final Grade"), title),
      x = ifelse(is.null(xlabel), xvar, xlabel),
      y = ylabel
    ) +
    theme(legend.position = "none")

  if (!is.null(fill_colors)) {
    p <- p + scale_fill_manual(values = fill_colors)
  }

  return(p)
}
```

```{r, warning=FALSE}
p1 <- make_boxplot(por_df, "school", c("GP" = "lavender", "MS" = "orange"), xlabel = "School")
p2 <- make_boxplot(por_df, "schoolsup", c("yes" = "purple", "no" = "tan"), xlabel = "School Support")
p3 <- make_boxplot(por_df, "Mjob", title = "Mother's Job vs Final Grade", xlabel = "Mother's Job")
p4 <- make_boxplot(por_df, "Fjob", title = "Father's Job vs Final Grade", xlabel = "Father's Job")
p5 <- make_boxplot(por_df, "internet", c("yes" = "forestgreen", "no" = "gray60"), xlabel = "Internet Access")
p6 <- make_boxplot(por_df, "sex", c("F" = "red", "M" = "steelblue"), xlabel = "Sex")

p1 | p2
p3 / p4
p5 | p6
```


The GP school has a higher median G3 among the two. School support indicates 
struggling students--its median is slightly lower than those not receiving support.

For parent professions, the 95% median line indicates teacher parent profession
is highest, and at home is lowest. Health is highest with teacher for mother's,
and in the middle range for father's.

Having internet access is significant as well, and so is the student's
sex: having internet and being female having 95% confidence medians
above the opposite groups.

```{r}
p_reason <- make_boxplot(por_df, "reason", 
                         title = "Reason for Choosing School vs Final Grade",
                         xlabel = "Reason")

p_higher <- make_boxplot(por_df, "higher", fill_colors = c("yes" = "green", "no" = "red"),
  title = "Pursuing Higher Education vs Final Grade",
  xlabel = "Plans for Higher Ed")

p_reason / p_higher
```

Both of these factors indicate motivation. So, as is evident, they have significantly different medians between each of these two sets of groups.

```{r}
ggplot(por_df, aes(failures, G3)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", color = "darkred") +  # se = TRUE is default
  labs(title = "Past Failures vs Final Grade", x = "Number of Past Failures", y = "G3 (Final Grade)") +
  theme_bw()
```

This is the strongest predictor that is not based on G2 grade. Its R^2 is
rather weak, but bt tracing a horizontal line from the left to the
right, finding if the left's shaded bottom is above a point on the right's, one finds:
- 0 failures has a significantly greater predicted final grade than the rest 
- 1 failure has a significantly greater predicted final grade than 3 failures.

### Research Question 2: How is alcohol consumption among Portuguese high school students associated with their academic performance and social activities? 

\section*{Grade Distributions by Alcohol Levels} 

**Final Grade by Workday Alcohol Category:**
```{r}
ggplot(por_df, aes(x = Dalc, y = G3)) +
  geom_boxplot(outlier.alpha = 0.4, fill = "lightcoral") +
  labs(
    title = "Distribution of Final Grades Across Workday Alcohol Levels",
    x = "Workday Alcohol Level",
    y = "Final Grade (G3)"
  ) +
  theme_minimal()
```

**Final Grade by Weekend Alcohol Category:**
```{r}
ggplot(por_df, aes(x = Walc, y = G3)) +
  geom_boxplot(outlier.alpha = 0.4, fill = "steelblue") +
  labs(
    title = "Distribution of Final Grades Across Weekend Alcohol Levels",
    x = "Weekend Alcohol Level",
    y = "Final Grade (G3)"
  ) +
  theme_minimal()
```
---

\section*{Social Activities vs Alcohol Consumption} 

**Going Out Frequency by Workday Alcohol:**
```{r}
ggplot(por_df, aes(x = Dalc, y = goout)) +
  geom_boxplot(outlier.alpha = 0.4, fill = "lightcoral") +
  labs(
    title = "Going Out with Friends Across Workday Alcohol Levels",
    x = "Workday Alcohol Level",
    y = "Going Out Frequency (1=Very Low, 5=Very High)"
  ) +
  theme_minimal()
```

**Free Time After School by Weekend Alcohol:**
```{r}
ggplot(por_df, aes(x = Walc, y = freetime)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "Free Time After School Across Weekend Alcohol Levels",
    x = "Weekend Alcohol Level",
    y = "Free Time (1=Very Low, 5=Very High)"
  ) +
  theme_minimal()
```




