---
title: 'Final Project Report'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 
author:
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
  pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
    include-in-header: preamble.tex
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE, warning = FALSE, error = FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

```{r, include=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(knitr)
library(MASS)  # For AIC/BIC 
library(car) # For vif
```

```{r, include = FALSE}
por_df <- read.csv("../data/student-por.csv", sep = ";")
head(por_df)

por_df <- por_df %>%
  mutate(
    school = as.factor(school),
    sex = as.factor(sex),
    address = as.factor(address),
    famsize = as.factor(famsize),
    Pstatus = as.factor(Pstatus),
    Mjob = as.factor(Mjob),
    Fjob = as.factor(Fjob),
    reason = as.factor(reason),
    guardian = as.factor(guardian),
    schoolsup = as.factor(schoolsup),
    famsup = as.factor(famsup),
    paid = as.factor(paid),
    activities = as.factor(activities),
    nursery = as.factor(nursery),
    higher = as.factor(higher),
    internet = as.factor(internet),
    romantic = as.factor(romantic),
    # ordered factor from 1-5 for very high, low etc.
    Dalc = factor(Dalc, levels = 1:5, ordered = TRUE), 
    Walc = factor(Walc, levels = 1:5, ordered = TRUE)
  )
```

:::: callout
::: {style="text-align: center"}
[**STUDENT NAME**]{style="color: blue;"}
:::

-   Xinyao Song (xinyaosong)
-   Matthew Chan (hchan837)
-   Darren Colianni (dcolianni)
-   Nic Chan (nicholaschan)
-   STUDENT 5 (NetID 5)
::::

# Abstract

This project investigates how various social, familial, and behavioral
factors are correlated with academic performance and average free times
among Portuguese high school students. Using the Student Performance
dataset compiled by Cortez and Silva in 2008, we explore three specific
research questions: the first two concerning the influence of parental
education and alcohol consumption on students’ final Portuguese grades
(G3), and the last one focusing on how one's relationship status
influence the amount of free time students report. From the result, we
can see that \[FINISH AT THE END\]

# Introduction

This project focuses on how different factors influence school
performance in Portuguese secondary school students. The dataset our
team used is the Student Performance dataset compiled by Cortez and
Silva in 2008. The dataset contains 649 observations and 30 features
collected through questionnaires from two Portuguese high schools, with
features covering aspects of family background, social environment,
academic records, and types of lifestyles. The primary objective of our
analysis is to examine how different social and behavioral factors
relate to the final Portuguese grade (G3), which is measured on a scale
from 0 to 20. By answering the three focused research questions below,
we aim to understand how various background and behavioral variables
correlate with or predict academic performance. The results from this
analysis could help inform educational strategies and potentially help
students with their stress and mental health.

## Research Question 1

After statistically controlling for weekly study time (on a 1–4 numeric
scale where 1 indicates less than 2 hours and 4 indicates more than 10
hours) and internet access (coded as “yes” or “no”), how is the father’s
highest education level (on a 0–4 numeric scale where 0 indicates no
formal education and 4 indicates higher education) associated with
Portuguese high school students’ final grades in Portuguese class (G3),
which are measured on a 0–20 numeric scale?

**Hypothesis:**

$H_0$: After controlling for weekly study time and internet access, the
father’s highest education level (0 = no formal education to 4 = higher
education) has no significant association with students’ final
Portuguese grade (G3, measured on a 0–20 scale).

$H_1$: After controlling for weekly study time and internet access, the
father’s highest education level (0 = no formal education to 4 = higher
education) is significantly positively associated with students’ final
Portuguese grade (G3, 0–20 scale).

## Research Question 2

Based on the number of alcoholic beverages consumed by Portuguese high
school students per weekday and weekend (on a 1-5 numeric scale where 1
indicates very low and 5 indicating very high), how does it affect
Portuguese high school students’ final grades in Portuguese class (G3),
which are measured on a 0–20 numeric scale?

**Hypothesis:**

$H_0$: There is no significant association between students’ alcohol
consumption levels during weekdays and weekends (1 = very low to 5 =
very high) and their final Portuguese grade (G3, 0–20 scale).

$H_1$: Students’ alcohol consumption levels during weekdays and weekends
(1 = very low to 5 = very high) are significantly associated with their
final Portuguese grade (G3, 0–20 scale).

## Research Question 3

Is there a statistically significant difference in reported average free
time (on a 1-5 categorical scale where 1 indicates very low and 5
indicates very high) between Portuguese high school students who are in
romantic relationships (coded as 1), and those who are not in romantic
relationships (coded as 0)?

**Hypothesis:**

$H_0$: There is no significant difference in average free time between
students in a romantic relationship and those who are not.

$H_1$: There is a significant difference in average free time (1 = very
low to 5 = very high) between students in a romantic relationship and
students who are not in romantic relationships.

# Data Processing

The Student Performance dataset does not contain any missing values or
fields with unclear meanings. Therefore, no data cleaning or
preprocessing was necessary for this project.

# Research Question 1

## Modeling Process

The aim is to observe if, after controlling for study time and internet
access, father's education significantly influenced the student's final
Portuguese grade. The question at hand calls for a partial significance
test, which works by having one one the models be a subset to another.
Therefore, if there is a significant difference, it is due to the
predictors added in between.\
\
The models were both linear and used the following predictors:

-   Model 1: Study time and internet access.

-   Model 2: Model 1's predictors as well as the education level of the
    student's father.

The two linear models were compared using an ANOVA test. Since the only
predictor added in model 2 was the father's education, if there was a
significant decrease in residual variance--as indicated by the resulting
p-value--it would be due to this predictor.

## Results

```{r, echo = FALSE}
anova_kable <- function(model1, model2) {
  aov_out <- anova(model1, model2)

  Model <- rownames(aov_out)
  Res.Df <- aov_out[, "Res.Df"]
  RSS <- signif(aov_out[, "RSS"], 5)
  Df <- ifelse(is.na(aov_out[, "Df"]), "", aov_out[, "Df"])
  Sum_of_Sq <- ifelse(is.na(aov_out[, "Sum of Sq"]), "", signif(aov_out[, "Sum of Sq"], 5))
  F <- ifelse(is.na(aov_out[, "F"]), "", signif(aov_out[, "F"], 4))

  # Extract p-values as numeric, then format with signif( ,2)
  pvals_raw <- aov_out$`Pr(>F)`
  pvals_signif <- ifelse(
    is.na(pvals_raw),
    "",
    formatC(signif(pvals_raw, 2), format = "e", digits = 2)
  )

  Signif <- cut(as.numeric(pvals_raw),
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                labels = c("***", "**", "*", ".", ""),
                right = TRUE)
  Signif[is.na(Signif)] <- ""

  df <- data.frame(
    Model = Model,
    Res.Df = Res.Df,
    RSS = RSS,
    Df = Df,
    `Sum of Sq` = Sum_of_Sq,
    F = F,
    `Pr(>F)` = pvals_signif,
    Signif = Signif,
    stringsAsFactors = FALSE
  )

  knitr::kable(df, align = "c",
               caption = "ANOVA Comparison Table")
}

lm_table <- function(model, digits = 2, pval_sci_threshold = 0.01, caption = "Linear Model Summary") {
  df <- broom::tidy(model) %>%
    dplyr::select(term, estimate, std.error, statistic, p.value) %>%
    dplyr::rename(
      Term = term,
      Estimate = estimate,
      `Std. Error` = std.error,
      `t value` = statistic,
      `Pr(>|t|)` = p.value
    )

  # Format values
  df$Estimate <- signif(df$Estimate, digits)
  df$`Std. Error` <- signif(df$`Std. Error`, digits)
  df$`t value` <- signif(df$`t value`, digits)

  # Format p-values with smart thresholding
  df$`Pr(>|t|)` <- ifelse(
    is.na(df$`Pr(>|t|)`),
    "",
    ifelse(
      df$`Pr(>|t|)` < pval_sci_threshold,
      formatC(signif(df$`Pr(>|t|)`, digits), format = "e", digits = digits),
      formatC(round(df$`Pr(>|t|)`, digits), format = "f", digits = digits)
    )
  )

  # Add significance stars
  pval_raw <- broom::tidy(model)$p.value
  df$Signif <- cut(pval_raw,
                   breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                   labels = c("***", "**", "*", ".", ""),
                   right = TRUE)
  df$Signif[is.na(df$Signif)] <- ""

  # Return styled kable
  knitr::kable(df, align = "c", caption = caption)
}
```

Here were the results:

```{r, echo=FALSE}
f1 <- G3 ~ studytime + internet
f2 <- G3 ~ studytime + internet + Fedu

model1 <- lm(data=por_df, formula = f1)
model2 <- lm(data=por_df, formula = f2)
anova_parsig <- anova(model1, model2)

anova_kable(model1, model2)
```

## Interpretation

With a p-value of approximately 2.30 \* $10^{-6}$, there is sufficient
evidence to reject the null hypothesis,\
$H_0$: that after controlling for weekly study time and internet access,
the father’s highest education level has no significant association with
students’ final Portuguese grade. Therefore, opting instead for the
alternative hypothesis, the father's highest education level does play a
role in the students' final Portuguese grade.

## Visualization and Communication (Model Checking)

Model checking was performed to ensure this result from the previous
section was accurate. To assess whether this model is appropriate, we
examined Model 2, which contains the same predictors as Model 1 in
addition to father's education. For the partial significance test to be
valid, Model 2 must satisfy the standard linear regression assumptions:
independence, no structure in residuals, normality of residuals, and
homoscedasticity (equal variance across fitted values).

Independence is satisfied because the dataset includes the full
population of students from each school during that year. Additionally,
because this is an observational study with no temporal or treatment
structure, residual independence is expected and visually confirmed. It
only remains to be seen whether the remaining conditions—normality and
equal variance—hold true.

```{r Functions, echo=FALSE}
shapiro_pval <- function(resids, digits=3){
  shap <- shapiro.test(resids)
  signif(shap$p.value, digits)
}

qqplott <- function(residuals){
  ggplot(data.frame(residuals), aes(sample = residuals)) +
    stat_qq() +
    stat_qq_line() +
    labs(title = "Q-Q Plot of Residuals", x = "Theoretical Quantiles", y = "Sample Quantiles") +
    theme_minimal()
}

hist_resids <- function(resids, bins = 30) {
  df <- data.frame(residuals = resids)
  ggplot(df, aes(x = residuals)) +
    geom_histogram(bins = bins, fill = "steelblue", color = "black", alpha = 0.7) +
    labs(title = "Student Grade Residuals", x = "Residuals", y = "Count") +
    theme_minimal()
}


residplot_fitval <- function(model) {
  df <- data.frame(Fitted = fitted(model), Residuals = residuals(model))
  ggplot(df, aes(x = Fitted, y = Residuals)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
    labs(title = "Residuals vs. Fitted Values", x = "Fitted Values", y = "Residuals") +
    theme_minimal()
}
```

```{r, echo=FALSE}
resids <- model2$residuals
```

### Normality of Residuals

```{r Normality Charts, echo=FALSE}
hist_resids(resids)
qqplott(resids)
```

```{r Calculations, echo=FALSE}
shap_test <- shapiro.test(resids)
shap_pval <- signif(shap_test$p.value, 3)
off_values_percent <- round(100 * (sum(resids < -7)/nrow(por_df)), 2)
```

The Q-Q plot indicates a deviation from normality in the residuals, and
likewise, a Shapiro-wilk test result of `r shap_pval`(p-value \< 0.05)
corroborates this deviation. Nevertheless, as evident through the
histogram, the deviation is primarily due to a relatively small group of
students who scored significantly lower than predicted, causing a mild
left skew. These values represent approximately `r off_values_percent`%
of the residuals. Aside from this small subset, the residuals are
normally distributed and centered around zero. \
\
Given the large sample size and that the deviation stems from a small
portion of the data, the normality assumption is considered reasonably
satisfied for inference purposes.

### Equal Variances Across Fitted Values

```{r Residplotchart, echo=FALSE}
residplot_fitval(model2)
```

The residuals appear to have approximately constant variance across the
range of fitted values, with no clear funneling or systematic spread.
Thus, there is no strong evidence to reject the homoscedasticity
assumption.

Hence, all assumptions for the partial significance test performed were
reasonably satisfied. Therefore, the previously found p-value of
approximately 2.30 \* $10^{-6}$ can be trusted.

## Conclusion and Recommendation

After controlling for study time and internet access, the father’s
education level remains a significant positive predictor of students’
final Portuguese grades. This likely reflects the academic support that
more educated parents are able to provide at home; eg, helping with
homework and explaining difficult topics. To support those lacking this
advantage, schools should consider expanding tutoring and afterschool
programs to foster a more equitable learning environment. Additionally,
making greater use of school libraries—through activities like book
clubs or guest author visits—could help motivate students to read more
and strengthen their Portuguese language skills.

# Research Question 2

## Modeling Process

## Results

## Interpretation

# Research Question 3

## Modeling Process

## Results

## Interpretation

# Conclusion
