---
title: 'Final Project - Step 2 (15 Points)'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author:
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```



:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Xinyao Song   (6194773)
- STUDENT 2    (NetID 2)
- STUDENT 3    (NetID 3)
- STUDENT 4    (NetID 4)
- STUDENT 5    (NetID 5)
:::


::: {.callout-caution}
## Due Date
The deadline for this step is **Friday, May 9, 2025**.
:::


:::{.callout-tip}
# Instructions
In this step, you will develop clear research questions and hypotheses based on your selected dataset, and conduct a thorough Exploratory Data Analysis (EDA). This foundational work is crucial for guiding your analysis in the following steps.

:::

# Step 2: Research Questions, Hypotheses, and Exploratory Data Analysis (EDA)


## Research Questions

**Question 1**  

- How does weekly study time, parents’ education level, and internet access influence a student’s average Portuguese grade (average of G1, G2, and G3)?

**Question 2**  

- How is alcohol consumption among Portuguese high school students associated with their academic performance and social activities? 

**Question 3**  

- Does the average free time differ between students who are in a romantic relationship and those who are not? 

## Hypotheses

**Hypothesis 1**  

- At least one of the predictors (weekly study time, parents’ education level, or internet access) has a significant effect on students’ average Portuguese grade. 

**Hypothesis 2**  

- Alcohol consumption is significantly associated with either academic performance or social activity levels among Portuguese high school students. 

**Hypothesis 3**  

- There is a significant difference in means between students in a relationship and those who are not.

## Exploratory Data Analysis (EDA)

## Data Cleaning
```{r}
library(dplyr)
library(tidyr)
library(readr)

# Preprocess student data frame
por_df <- readr::read_delim("../data/student-por.csv", delim = ";")

por_df <- por_df %>%
  mutate(
    school = as.factor(school),
    sex = as.factor(sex),
    address = as.factor(address),
    famsize = as.factor(famsize),
    Pstatus = as.factor(Pstatus),
    Mjob = as.factor(Mjob),
    Fjob = as.factor(Fjob),
    reason = as.factor(reason),
    guardian = as.factor(guardian),
    schoolsup = as.factor(schoolsup),
    famsup = as.factor(famsup),
    paid = as.factor(paid),
    activities = as.factor(activities),
    nursery = as.factor(nursery),
    higher = as.factor(higher),
    internet = as.factor(internet),
    romantic = as.factor(romantic)
  )

# Check for missing values
print(anyNA(por_df))
```
From the missing value check above, we can see that there are no missing values in this dataset. Therefore, no data cleaning is necessary.

## Descriptive Statistics
```{r}
summary(por_df)
```



## Data Visualization

